\documentclass[11pt]{book}

\input{worksheets-preamble.tex}
\begin{document}
%
%
\typeout{************************************************}
\typeout{Worksheet  Activity: Constructible Numbers}
\typeout{************************************************}
%
\newgeometry{left=1in, right=1in, top=1in, bottom=1.25cm}
\begin{worksheet-subsection}{Activity: Constructible Numbers}{}{Activity: Constructible Numbers}{}{}{x:worksheet:ws_fields-constructable-numbers}
\begin{introduction}{}%
We have considered what geometric shapes we could or could not construct.  We are left with three big questions: is it possible to \emph{trisect an angle}, to \emph{double a cube}, or to \emph{square a circle}.  To answer these questions, we must ``algebratize'' geometric constructions.%
\par
Start with two constructible points \(0\) and \(1\) \emph{one unit} apart.  We define \terminology{constructible} recursively from this base case:%
\begin{enumerate}[label=(\alph*)]
\item{}A \terminology{constructible line} is a line passing through two constructible points.%
\item{}A \terminology{constructible circle} is a circle whose radius is a constructible number and whose center is a constructible point.%
\item{}A \terminology{constructible point} is the intersection of two constructible lines, two constructible circles, or a constructible line and a constructible circle.%
\end{enumerate}
%
\par
We say a number \(a\) is \terminology{constructible} provided \(a = 0\) or there are two constructible points distance \(|a|\) apart.  So far we have constructible numbers 0, 1, and \(-1\).  What else is constructible?%
\par
For this activity, use GeoGebra.  Use only the ``new point'' tool (to place points at the intersections of lines, circles, or both), the ``line through two points'' tool, and the ``compass'' tool (under the circle menu).  You can also use the arrow to drag things around if you need to.%
\end{introduction}%
\begin{divisionexercise}{1}{}{}{p:exercise:NeE}%
Show that the numbers 2 and 4 are constructible.  Then show that the number \(3 = 4 - 1\) is constructible.%
\end{divisionexercise}%
\begin{divisionexercise}{2}{}{}{p:exercise:tlN}%
If \(a\) and \(b\) are constructible numbers, are the numbers \(a+b\) and \(a-b\) also constructible?%
\end{divisionexercise}%
\begin{divisionexercise}{3}{}{}{p:exercise:ZsW}%
Suppose \(a\) and \(b\) are constructible.  Construct a triangle containing a base of unit length adjacent to a side of length \(a\).  Construct a similar triangle with where the side corresponding the unit length side now has length \(b\).  What is the length of the side corresponding to the \(a\)-length side?%
\begin{image}{0.375}{0.25}{0.375}%
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}
  {\tiny
  \draw (0,0) -- (1.5,0) node[midway, above]{1};
  \draw (0,0) -- (1,1) node[midway,below]{$a$} -- (1.5,0);
  \draw (0,0) -- (2.25,0) node[midway,below]{$b$};
  \draw (0,0) -- (1.5,1.5) node[midway,above]{?} -- (2.25,0);
  }
\end{tikzpicture}
}%
\end{image}%
\end{divisionexercise}%
\begin{divisionexercise}{4}{}{}{p:exercise:FAf}%
Explain how you can modify the above construction to prove that if \(a\) and \(b\) are constructible, then \(a/b\) is constructible.%
\end{divisionexercise}%
\begin{divisionexercise}{5}{}{}{p:exercise:lHo}%
Given constructible number \(c\), explain how you can construct the figure below.  The vertical line should be perpendicular to the horizontal line, which is the diameter of the circle.%
\begin{image}{0.35}{0.3}{0.35}%
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1.0cm,y=1.0cm, scale=0.75]
\draw (-1.,0.)-- (5.,0.);
\draw [shift={(2.,0.)}] plot[domain=0.:3.141592653589793,variable=\t]({1.*3.*cos(\t r)+0.*3.*sin(\t r)},{0.*3.*cos(\t r)+1.*3.*sin(\t r)});
\draw (0.18,2.3848689691469427)-- (0.18,0.);
\begin{scriptsize}
\draw [fill=black] (-1.,0.) circle (1.5pt);
\draw (-1, 0) node[below] {$0$};
\draw [fill=black] (5.,0.) circle (1.5pt);
\draw (5,0) node[below] {$c$};
\draw [fill=black] (0.18,2.3848689691469427) circle (1.5pt);
\draw [fill=black] (0.18,0.) circle (1.5pt);
\draw (0.18,0.) node[below] {$1$};
\end{scriptsize}
\end{tikzpicture}
}%
\end{image}%
What is the length of the vertical line?%
\end{divisionexercise}%
\begin{divisionexercise}{6}{}{}{p:exercise:ROx}%
Let \(\mathfrak C\) be the set of all constructible numbers.  What sort of set is this?  Is it a group?  A ring?  A field?  Is it one of these we know about already?%
\end{divisionexercise}%
\end{worksheet-subsection}
\restoregeometry
\end{document}
