\documentclass[11pt]{book}

\input{worksheets-preamble.tex}
\begin{document}
%
%
\typeout{************************************************}
\typeout{Worksheet  Activity: (Sub)Normal Series}
\typeout{************************************************}
%
\newgeometry{left=1in, right=1in, top=1in, bottom=1.25cm}
\begin{worksheet-subsection}{Activity: (Sub)Normal Series}{}{Activity: (Sub)Normal Series}{}{}{x:worksheet:ws_grp-struct-series}
\begin{introduction}{}%
Given a group, we can look at subgroups.  We say that a sequence of subgroups%
\begin{equation*}
G = H_n \supset H_{n-1} \supset \cdots \supset H_1 \supset H_0 = \{e\}
\end{equation*}
is a \terminology{subnormal series} provided each \(H_i\) is normal in \(H_{i+1}\), and a \terminology{normal series} if each \(H_i\) is normal in \(G\).%
\par
A non-trivial group \(G\) is called \terminology{simple} provided it has no non-trivial normal subgroups.  We say that a subnormal series is a \terminology{composition series} and that a normal series is a \terminology{principle series} if every quotient group \(H_{i+1}/H_i\) is simple.%
\end{introduction}%
\begin{divisionexercise}{1}{}{0.15}{p:exercise:bSq}%
Find a subnormal series for \(D_4\). Is it a normal series?%
\end{divisionexercise}%
\begin{divisionexercise}{2}{}{0.15}{p:exercise:HZz}%
Find two different normal series for \(\Z_{60}\) of length 3 (length is the number of proper inclusions).%
\end{divisionexercise}%
\begin{divisionexercise}{3}{}{0.15}{p:exercise:ogI}%
Find the quotient groups \(H_{i+1}/H_i\) for both series above.  How are these related?  Are the series composition series?%
\end{divisionexercise}%
\begin{divisionexercise}{4}{}{0.01}{p:exercise:UnR}%
Find a composition series for \(\Z_{60}\).  Can you take it to be a \terminology{refinement} of the normal series you found above?%
\end{divisionexercise}%
\end{worksheet-subsection}
\restoregeometry
\end{document}
